---
title: "1.2 Programming in R - Tutorial"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)
```

---

At the end of this tutorial you should be able to:

* Understand how to execute code in an R Markdown file
* Identify output of executed code
* Write comments
* Perform basic math in R

---

*IMPORTANT NOTE*
You will not receive automatic feedback on this tutorial. If you are unsure of how to write any of the tutorial commands, please refer to the lecture supplement. All subsequent tutorials provided in R Markdown will have automatic feedback provided by Otter Grader.

---

## 1.2.2 R Markdown

---

Execute the following line of code and identify the output:

```{r}
5 - 4
```

Turn the following line of code into a comment and execute the code. Is there any output? Why or why not?

```{r}
4 + 2
```

The following line of code takes the number of employees at a large research facility at the beginning of 2008, subtracts the number of employees that have left, and adds the number of new hires throughout the year. Add a comment to the blank line above the calculation to explain what the calculation is to someone else who is looking at the code. 

```{r}

1039 - 34 + 67
```

Write a line of code in the empty code block that takes the number of employees at the research facility at the beginning of 2009 (the number output above) and calculates the number of employees at the beginning of 2010. During 2009, 27 employees left, and there were 32 new hires.

```{r}

```

---

## 1.2.3 Math in R

---

You have a 4 M solution of NaOH. You want to make 150 mL of a 2 M solution. 

Use the following formula to calculate the volume of the initial solution required: 

V1 = (C2 x V2)/C1

Where:
- C1 is the initial concentration (M)
- V1 is the initial volume (mL)
- C2 is the final concentration (M)
- V2 is the final volume (mL)

From this point, all questions in tutorials will be followed by a code block containing the comment "# Write and execute your code here" to indicate that you should write code in the code block below the comment, like this:

```{r}
# Write and execute your code here

```

Calculate the number of moles in the resulting 2 M NaOH solution. 

n = M × V

Where:
- M is molarity (mol/L)
- V is volume (L)

Note the volume is in litres (L) in the formula above, but the volume you need to use is in mL (150 mL), so you will need to include a conversion in your code (150 / 1000).

```{r}
# Write and execute your code here

```

NaOH has a molecular weight of 39.997 Daltons (g/mol). Calculate the grams of NaOH in the resulting solution using the number of moles you calculated in the previous question.

m = n x w

Where:
- n is moles (mol)
- w is molecular weight (g/mol)

```{r}
# Write and execute your code here

```

Calculate the number of atoms in the resulting solution. (n = result of an earlier calculation (moles)).

a = n x Avogadro's number

Where:
- n is moles (mol)
- Avogadro's number is 6.022 x 10^23 (atoms/mol)

Hint: n is the result of an earlier calculation (mol).

```{r}
# Write and execute your code here

```

---

What will be the output of the following equation if it is typed into R? Think of the answer before testing in the code block below.

20 - 14 * 2^3 + 2

a) 50
b) -90
c) 60

```{r}
# Write and execute your code here

```

Add brackets to the equation below to make the output equal 9:

```{r}
# Write and execute your code here
5 * 2 + 4 / 3 - 1
```

Change the brackets to now make the output equal 12:

```{r}
# Write and execute your code here
5 * 2 + 4 / 3 - 1
```

---
