---
title: "Week 1 Assignment"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)

library(ottr)
library(testthat)
```
**Run the above block to load required packages**

---

## Overview

---

This assignment covers the content in week 1 lectures and tutorials. 

#### Mark breakdown
- Question 1 (4 marks)
- Question 2 (6 marks)
- Question 4 (6 marks)

Note that when a question asks you to assign a value to a variable, you must use the exact variable name to get the marks for the question. The question will provide the variable name in quotation marks, although the variable in your code should not have quotation marks around it. For example, view the following and question and answer:

Assign the result of the calculation 1 + 1 to a variable named "q.9".

```{r}
# Write and execute your code here
q.9 <- 1 + 1
```

---

## Background

---

The *doubling time* of a bacterial species is the amount of time it takes for the population of the bacteria to double in size.

The doubling time of Staphylococcus aureus (S. aureus) grown under *lab conditions* is: 0.4 hours

The doubling time of Staphylococcus aureus (S. aureus) grown *in the wild* is: 1.87 hours

---

## QUESTIONS

---

### Question 1 

#### a.
(1 mark)
Assign the name of the bacterial species, Staphylococcus aureus, to a variable named "bacterial_species".

```{r Assignment.W1.1.a}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.1.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(1 mark)
Update the variable `bacterial_species` so that it contains the shorter version of the species name: S. aureus.

```{r Assignment.W1.1.b}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.1.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c.
(1 mark)
Assign the doubling time of S. aureus under lab conditions to a variable named "S.aureus_lab_doubling_time".

```{r Assignment.W1.1.c}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.1.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d.
(1 mark)
Assign the doubling time of S. aureus in the wild to a variable named "S.aureus_wild_doubling_time".

```{r Assignment.W1.1.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.1.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---

## Question 2

The formula for population growth based on doubling time is:

N = No x 2 ^ (t/d)

Where:
- N is the size of the final population
- No is the size of the initial population
- t is the time (in hours)
- d is the doubling time (in hours)

#### a.
(2 marks)
A colony of S. aureus with a population of 100 is grown in the *lab*. What is the population size after 2 hours? Assign the final population size to a variable named "S.aureus_lab_population".

```{r Assignment.W1.2.a}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.2.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(2 marks)
A colony of S. aureus with a population of 100 is grown in the *wild*. What is the population size after 2 hours? Assign the final population size to a variable named "S.aureus_wild_population".

```{r Assignment.W1.2.b}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.2.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c. 
(2 marks)
Using the results from 2.a and 2.b, how many *times more* S. aureus are there in the lab grown population compared to the wild grown population after 2 hours of growth? Assign the result of the calculation to a variable named "S.aureus_lab_over_wild".

```{r Assignment.W1.2.c}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.2.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---

## Question 3

Assign a logical value to the specified variable name to indicate if the statement is true or false.

#### a. 
(1 mark)
True or False: Character type values cannot contain numbers. Answer by assigning a logical value to a variable called "q.3.a".

```{r Assignment.W1.3.a}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.3.a.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### b.
(1 mark)
True or False: If the following piece of code was executed it would throw an error:

    sample_description <- "S. aureus was grown on the plate labelled "SA sample 3""
    
Answer by assigning a logical value to a variable called "q.3.b".

```{r Assignment.W1.3.b}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.3.b.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### c. 
(1 mark)
True or False: It doesn't matter whether you use capital letters when you type out a logical value. Answer by assigning a logical value to a variable called "q.3.c".

```{r Assignment.W1.3.c}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.3.c.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### d. 
(1 mark)
True or False: You can include the letter e in numeric values. Answer by assigning a logical value to a variable called "q.3.d".

```{r Assignment.W1.3.d}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.3.d.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### e. 
(1 mark)
True or False: When choosing a variable name, it is better to use names like "bacterial_species" and "S.aureus_lab_doubling_time" instead of "b_sp" and "S.a_lab_time". Answer by assigning a logical value to a variable called "q.3.e".

```{r Assignment.W1.3.e}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.3.e.R")
```
**IGNORE the above block: test results will not appear in assignments** 

#### f.
(1 mark)
True or False: If the following line of code was executed it would throw an error:

    1st.name <- "James"
    
Answer by assigning a logical value to a variable called "q.3.f".

```{r Assignment.W1.3.f}
# Write and execute your code here

```

```{r}
. = ottr::check("tests/Assignment.W1.3.f.R")
```
**IGNORE the above block: test results will not appear in assignments** 

---
