---
title: "1.3 Data Types & Variables"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = TRUE)
```

---

## 1.3.1 Data Types (Character)

---

### Data Type Overview

A *data type* is an attribute that classifies a piece of data. Different data types can have different values.

There are 3 main data types in R:

- Character type	(used for text)
- Numeric type	(used for numbers)
- Logical type	(used for true and false values)

There are three other types, but they are used rarely and we will not be learning them in this course.

Different operations can be performed on different data types. For example, in the the previous section we used R to perform calculations using numbers (numeric type). Numeric type values can be added together, however as we cannot perform mathematical operations with words, we could not add together two character type values.

Each data type has its own uses and syntax, which will be explored individually.

---

### Character Type

Character type values are text values. Words, phrases, and sentences in R are character type.

Character type values can contain:
- Letters
- Numbers
- Spaces
- Punctuation marks
- Other symbols on your keyboard: $, %, #, *, ~, etc.

Character type values must always be enclosed in double quotation marks "". 

Let's look at some examples of character type values. Here is the word hello as a character type value.

```{r}
"hello"
```

What would happen if we had typed hello without quotation marks?

```{r}
hello
```

The code *throws* an *error*. Errors are "thrown" whenever R cannot interpret and execute the code. When an error is thrown, an *error message* is also output. In this error message it states that the object 'hello' is not found, it saw the word hello and did not know how to interpret it.

R code is read and executed by the R *interpreter*, a program that reads and executes code written in a programming language (in this case, R). When the interpreter cannot execute the code it throws an error. When hello is executed without quotation marks, the interpreter does not know what the word represents and throws an error.

What if we want to create a character type value that itself contains quotation marks? For example: Sample label: "EGH1"

Before you execute the code below, notice the red x to the left of the line number. This is a warning that there is something wrong with the code written, as is the jagged red line below EGH1. If you hover over either of these, a message will pop up: unexpected token 'EGH1'

Let's run it to see what happens.

```{r}
"Sample label: "EGH1""
```

In this error message it states it saw an "unexpected symbol", in this case it saw the word EGH1 and did not know how to interpret it. 

If we look at the back at the code, we can see that because of the use of double quotation marks two times, there are actually two character type values:
  "Sample label: " and "" with the uninterpretable word EGH1 in the middle.
In order for R to interpret letters as a character value, they must be within quotation marks.

Instead we can use single quotations within the double quotations of the character value:

```{r}
"Sample label: 'EGH1'"
```

Character values are used for all text values in R. Here are a few examples of character value uses in bioinformatic analysis:

- Sample descriptions   `“EGH1 deletion mutant”`
- Cell types            `“HEK293”`
- Gene names            `“TP53”`
- Patient cancer type   `“pancreatic, stage III”`

---

## 1.3.2 Data Types (Numeric & Logical)

---

### Numeric Type

Numeric type values (also referred to as "double" type) are real numbers. 

Numeric type values can contain:
- Numbers: 0-9
- A single period to indicate a decimal point
- A single hyphen at the front to indicate a negative number
- The letter “e” to indicate “10 to the power of”

This is the type of value we were using in the math portion of the 1.2 module. Let's look at a few more examples.

```{r}
1067
```

Floating point numbers.

```{r}
99.99999
```

Negative numbers.

```{r}
-193830
```

Scientific e notation. (This is Avogadro's number: 6.022 x 10^23)

```{r}
6.022e+23
```

What would happen if we put a number inside quotation marks? 

```{r}
"-193830"
```

Is it still numeric type? No it is character type. This can be seen in the ouput as the value is contained in double quotation marks. Remember that character type values can contain numbers, and all values put inside quotation marks are interpreted as character type values.

```{r}
"-193830" + 1067
```

As was mentioned earlier, different operations can be performed on different data types. Addition is an operation that can be performed on numeric type values, but cannot be performed on character type values.

Although the code in the previous code chunk throws an error, there were no warnings from RStudio that there is a mistake in the code. RStudio will not catch every error before it happens.

Numeric values have numerous applications in bioinformatic analysis. Any number resulting from an analysis will be used as a numeric value in R. Here are a few examples:

- Concentrations            `10 # mol/L`
- Population growth rate    `0.237 # per year`
- Gene expression level     `39.01 # TPM`
- Patient age               `89 # years old`

Note that the units of each numeric type value were shown as a comment after the value. Units cannot be stored within a value, however a comment after the value with the units can be useful reminder when reviewing code.

---

### Logical Type

Logical type values (sometimes referred to as "Boolean" values or "Booleans") are used to indicate if something is true or false, and therefore there are only two logical type values:

TRUE
FALSE

They must be written in capital letters, like this:

```{r}
TRUE
```

```{r}
FALSE
```

What happens if we do not write the values in capital letters?

```{r}
True
```

Another error message. This one is telling us that R cannot inpret the value True. R cannot interpret True, true, False, or false. It can only interpret the values as capitalized TRUE and FALSE. In this course, code will always been shown with logical values `TRUE` or `FALSE`, however, there is one other R interpretable representation of these values: `T` and `F`:

```{r}
T
```

```{r}
F
```

Note, however, that the output of `T` and `F` are `TRUE` and `FALSE`.

Logical type values may not seem particularly useful now, but we will learn more about their uses next week in modules 2.2 and 2.3. Logical values are used in binary situations. Some examples of binary situations you may encounter in bioinformatics include:

- Mouse sex (male/female)
- Test positivity (pass/fail)
- Protein interactions (interact/don't interact)
- Patient in remission (yes/no)

---

## 1.3.3 Variables

---

### Creating Variables

*Variables* in R can be used to store values of any type. Let's start by looking at an example. 

To calculate the number of atoms are in 6 moles of water, Avogadro's constant (6.022×10^23 atoms/mol) is used:

atoms = moles * Avogadro's constant

The calculation can be performed in R using numeric values:

```{r}
6 * 6.022e+23
```

Now imagine you need to find the number of atoms in each of the following:

2.25 moles of NaOH
10 moles of KCl
4.3 moles of Ni

Avogadro's constant will need to be used for each calculation. Each time you type out the number you are more likely to make a mistake. Also, if you do not have it memorized, you may have to spend time double checking the value each time you need to use it. 

Instead, you can *assign* the numeric value of Avogadro's constant to a variable and use the variable as a value.

When a value is assigned to a variable in R, it becomes part of the *R environment*. The R environment keeps track of objects, including variables. When R sees the variable, it can access the value stored in it from the environment. In other words, the variable can be *interpreted* by R.

The assignment operator in R is a less than sign follwed by a hyphen: `<-`. The syntax for assigning a value to a variable is:

    variable <- value

Let's create a variable called "avogadros_constant" and assign it the numeric value 6.022e+23: 

```{r}
avogadros_constant <- 6.022e+23
```

There was no output when we ran this code chunk. When a variable is assigned to a value, there will be no output. To see the value of the variable, we can type the variable name and execute the code. This will output the value that is *stored* in the variable.

Let's look at what happens when we output `avogadros_constant`.

```{r}
avogadros_constant
```

The variable `avogadros_constant` is now equivalent to the numeric value 6.022e+23. Any time `avogadros_constant` is used, R interprets it as the value 6.022e+23.

---

### Using Variables

Let's find the number of atoms in 6 moles of water again. Instead of typing out Avogadro's constant, we will replace the numeric value with the variable `avogadros_constant`.

```{r}
6 * avogadros_constant
```

We can use the variable again to calculate the number of atoms in 2.25 moles of NaOH.

```{r}
2.25 * avogadros_constant
```

And to calculate the number of atoms in 10 moles of KCl. 

```{r}
10 * avogadros_constant
```

We can also assign the result of the calculation to a new variable. This way the result can be used in later calculations.

```{r}
atoms_in_Ni <- 4.3 * avogadros_constant
atoms_in_Ni
```

Notice that we put a second line of code directly after the assignment to output the value that was stored in the variable.

---

### Tab Completion

Any data type can be assigned to a variable. For example, the character type value "PIDD0000012" is assigned to a variable named `patient_ID` below.

```{r}
patient_ID <- "PID0000012"
patient_ID
```

To keep track of whether the patient has a low iron level, a logical value can be assigned to a variable named `patient_iron_low`.

```{r}
patient_iron_low <- FALSE
patient_iron_low
```

A useful feature of RStudio is *tab completion*. In the code chunk below are the letters "avog". Place your cursor after the letter "g" in the code chunk. Press tab. If you previously ran the code chunk that assigned a value to the variable `avogadros_constant`, the rest of the variable name should automatically fill in. RStudio will search for variables that start with the letters and complete the name for you.

```{r}
avog
```

If two variables start with the same few letters, like `patient_ID` and `patient_iron_low`, pressing tab will bring up a menu of variables that start with those letters. You can click the name of the variable you want to use to fill in the rest of the variable name.

```{r}
pati
```

---

## 1.3.4 Variable Names

---

### Variable Names

You might have noticed two things about variable names in the previous examples:

  1. Although they contain letters, like character type values, variable names do not have quotation marks around them.
  2. Variable names do not contain spaces.
  
Variable names cannot have quotation marks or spaces, they must only include:

- letters
- numbers
- underscores
- periods

Variable names *cannot*:
- Contain spaces or any special characters
- Start with a number, underscore, or a period followed by a number

Because of this, it is best to always start your variable name with a letter so that you know it will be valid.

Here is an example of invalid variable name.

```{r}
2021_number_of_new_patients <- 129
```

If you use an invalid variable name, R will throw an error to let you know, and depending on what mistake you are making it will warn you about it before execution, as above.

Instead, you could name the variable:

- number_of_new_patients_2021
- num_new_patients_2021
- NewPatientsCount2021
- new.patients.2021

There are many other valid names, these are just a few examples.

Note that when we use multiple words in our variable name, we can improve readability by adding a *delimiter* between each word because spaces cannot be used.

Underscore delimiter: `num_new_patients_2021`
Period delimiter: `num.new.patients.2021`

Or we can use uppercase letters to differentiate between words:

Camel case: `numNewPatients2021`
Upper camel case: `NumNewPatients2021`

It is important to be descriptive when creating variable names so that we can remember later what the value represents, or, so that someone else can understand your code. 

Note that when you looked the variable names above, even though you were never told what it was supposed to represent, you probably realized that the value stored in the variable represents that there were 129 new patients in 2021.

Here are a few examples of names for the variable that would not have been easy to figure out:

- np2021
- new_p_2021
- n_patient_21
- patients_21
- newpat21

Although these are bad names for human readability, they are not invalid variable names (no spaces or special characters, and they all start with a letter). So R would not throw an error if you used one of these names. Choosing a good variable name is important for the user.

---

### The RStudio Environment Tab

At this point you may have noticed some data popping up to the right of this markdown file under the "Environment" tab in the top right corner of the RStudio interface. Whenever a new variable is created, the variable name the value are added to the "Global Environment". If you have run all the code chunks in this file so far you should see the following variables along with their values in the Environment tab:

    atoms_in_Ni             2.58946e+24
    avogadros_constant      6.022e+23
    patient_ID              "PID0000012"
    patient_iron_low        FALSE
    
You can save your environment, also called your *workspace* using the save icon in the environment tab, and you can load a saved environment using the icon to the left of the save button. If you want to clear (i.e. delete) all the variables in your environment, use the button with the broom icon.

---

### Overwriting Existing Variables

Earlier we assigned the logical value `FALSE` to the variable `patient_iron_low`.

```{r}
patient_iron_low 
```

If the patient returns to the doctor for a blood test and the doctor finds that the patient now has low iron levels, the variable can be *overwritten*, or in other words, edited.

We can overwrite the value of `patient_iron_low` from `FALSE` to `TRUE` by simply assigning the value `TRUE` to the same variable name.

```{r}
patient_iron_low <- TRUE
patient_iron_low
```

Now when you execute the variable, it outputs `TRUE`. 

You can also look at the "Environment" to the right as see that the value of `patient_iron_low` is now `TRUE`.

        patient_iron_low        TRUE

Overwriting variables is very useful when you want to run the same code, but update one number. However, it can also cause big problems if you do it by accident. It is good practice to check that your variables contain the expected values.

---
